---
title: "Calculating Urbanicity Scores for Reference Set"
author: "Lev Kolinski"
date: "`r Sys.Date()`"
output: html_document
---
## Setting File Paths and Arguments
Update the file paths/objects below with your relevant information. All arguments will be fed directly into the compute_uranicity_iterative() call below.
```{r set_objects}
Sys.setenv(RETICULATE_PYTHON = "C:/Users/tmbar/AppData/Local/r-miniconda/envs/rgee/python.exe")

# Set Path To Local Overpass Instance
osmdata::set_overpass_url("http://localhost:12345/api/interpreter")

# where your reference coordinate set is stored
reference_coordinate_set_fp <- "C:/Users/tmbar/OneDrive - Duke University/Active Projects/pecahn/ref_pts.csv"

# your Google Earth Engine project name
ee_project_name <- "ee-tmbarrett2"

# year/years to calculate population density and nighttime light intensity
year_or_years <- c(2015:2025)

# file path where your urbanicity batch results should be saved
output_fp <- "C:/Users/tmbar/OneDrive - Duke University/Active Projects/pecahn/ref_pts_results"

# uncomment and run if needed to authenticate your Google Earth Engine API
#ee_Authenticate(auth_mode = "localhost")

```

## Loading Packages
```{r setup}
library(pecahnurbanicity)
library(tidyverse)
library(reticulate)
library(rgee)
```

## Writing a function to batch and save results 
Calls in chunks of 5 rows, saving intermediate results so an OSM failure mid-loop doesn't wipe your progress. 

- Creates 1 km local and 5km regional bounding boxes
- Computes "all" metrics

```{r compute_urbanicity_batched function}
compute_urbanicity_batched <- function(df, ee_project, year_or_years, 
                                        batch_size = 5, 
                                        save_prefix = NULL) {
  n <- nrow(df)
  batch_starts <- seq(1, n, by = batch_size)
  all_results <- list()
  
  for (i in seq_along(batch_starts)) {
    start_idx <- batch_starts[i]
    end_idx   <- min(start_idx + batch_size - 1, n)
    
    cat(sprintf("[Batch %d/%d] Rows %d–%d of %d\n", 
                i, length(batch_starts), start_idx, end_idx, n))
    
    df_batch <- df[start_idx:end_idx, ]
    
    tryCatch({
      bbox_1km_batch <- create_bounding_boxes(df_batch, distance_km = 1)
      bbox_5km_batch <- create_bounding_boxes(df_batch, distance_km = 5)
      
      result_batch <- compute_urbanicity_iterative(
        local_bboxes          = bbox_1km_batch,
        regional_bboxes       = bbox_5km_batch,
        metrics               = "all",
        ee_project            = ee_project,
        search_buffer         = 1,
        population_years      = year_or_years,
        nighttime_light_years = year_or_years,
        verbose               = TRUE
      )
      
      all_results[[i]] <- result_batch
      
      # Optional: save each batch to disk immediately as insurance
      if (!is.null(save_prefix)) {
        saveRDS(result_batch, 
                file = sprintf("%s_batch_%03d.rds", save_prefix, i))
        cat(sprintf("  -> Saved batch %d to disk\n", i))
      }
      
    }, error = function(e) {
      cat(sprintf("  [ERROR] Batch %d failed: %s\n", i, conditionMessage(e)))
      # Store NULL with a note so we know which batch failed
      all_results[[i]] <<- NULL  
    })
  }
  
  # Bind all successful batches
  successful <- Filter(Negate(is.null), all_results)
  if (length(successful) == 0) {
    warning("All batches failed — returning NULL")
    return(NULL)
  }
  
  bound <- bind_rows(successful)
  cat(sprintf("Done. %d/%d batches succeeded (%d rows total).\n",
              length(successful), length(batch_starts), nrow(bound)))
  return(bound)
}
```

## Getting Google Earth Engine ready
Update code as needed for your machine/project. This is the code I use for my device (Apple)

```{r GE_Engine}
ee <- reticulate::import("ee")
ee$Initialize(project = paste0(ee_project_name))
```

## Wrangling reference coordinate set to match expected input for compute_urbanicity_iterative()
```{r wrangling}

reference_coordinate_set <- read.csv(reference_coordinate_set_fp)

reference_coordinate_set <- reference_coordinate_set |> 
  rename(project = Classified) |> 
  mutate(community = paste0(lat, ", ", lon)) |> 
  mutate(project = case_when(
    project == 11 ~ "Very Low Density Grid Cell",
    project == 12 ~ "Low Density Rural Grid Cell",
    project == 13 ~ "Rural Cluster",
    project == 21 ~ "Suburban Grid Cell",
    project == 22 ~ "Semi-Dense Urban Cluster",
    project == 23 ~ "Dense Urban Cluster",
    project == 30 ~ "Urban Center",
    .default = NA
  ))

reference_coordinate_set_1<- reference_coordinate_set |> 
  filter(project=="Very Low Density Grid Cell") |> 
  select(project, community, lat, lon)

reference_coordinate_set_2<- reference_coordinate_set |> 
  filter(project=="Low Density Rural Grid Cell") |> 
  select(project, community, lat, lon)

reference_coordinate_set_3<- reference_coordinate_set |> 
  filter(project=="Rural Cluster") |> 
  select(project, community, lat, lon)

reference_coordinate_set_4<- reference_coordinate_set |> 
  filter(project=="Suburban Grid Cell") |> 
  select(project, community, lat, lon)

reference_coordinate_set_5<- reference_coordinate_set |> 
  filter(project=="Semi-Dense Urban Cluster") |> 
  select(project, community, lat, lon)

reference_coordinate_set_6<- reference_coordinate_set |> 
  filter(project=="Dense Urban Cluster") |> 
  select(project, community, lat, lon)

reference_coordinate_set_7<- reference_coordinate_set |> 
  filter(project=="Urban Center") |> 
  select(project, community, lat, lon)

```

## Running urbanicity metrics: Very Low Density Grid Cell
```{r compute_urbanictiy_1}
results_1 <- compute_urbanicity_batched(
  df          = reference_coordinate_set_1,
  ee_project  = ee_project_name,
  year_or_years = year_or_years,
  batch_size  = 5,
  save_prefix = paste0(output_fp, "subset1") 
)

```


## Running urbanicity metrics: Low Density Rural Grid Cell
```{r compute_urbanictiy_2}
results_2 <- compute_urbanicity_batched(
  df          = reference_coordinate_set_2,
  ee_project  = ee_project_name,
  year_or_years = year_or_years,
  batch_size  = 5,
  save_prefix = paste0(output_fp, "subset2") 
)

```


## Running urbanicity metrics: Rural Cluster
```{r compute_urbanictiy_3}
results_3 <- compute_urbanicity_batched(
  df          = reference_coordinate_set_3,
  ee_project  = ee_project_name,
  year_or_years = year_or_years,
  batch_size  = 5,
  save_prefix = paste0(output_fp, "subset3") 
)
```


## Running urbanicity metrics: Suburban Grid Cell
```{r compute_urbanictiy_4}
results_4 <- compute_urbanicity_batched(
  df          = reference_coordinate_set_4,
  ee_project  = ee_project_name,
  year_or_years = year_or_years,
  batch_size  = 5,
  save_prefix = paste0(output_fp, "subset4") 
)

```


## Running urbanicity metrics: Semi-Dense Urban Cluster
```{r compute_urbanictiy_5}
results_5 <- compute_urbanicity_batched(
  df          = reference_coordinate_set_5,
  ee_project  = ee_project_name,
  year_or_years = year_or_years,
  batch_size  = 5,
  save_prefix = paste0(output_fp, "subset5")
)
```

## Running urbanicity metrics: Dense Urban Center
```{r compute_urbanictiy_6}
results_6 <- compute_urbanicity_batched(
  df          = reference_coordinate_set_6,
  ee_project  = ee_project_name,
  year_or_years = year_or_years,
  batch_size  = 5,
  save_prefix = paste0(output_fp, "subset6") 
)

```


## Running urbanicity metrics: Urban Center
```{r compute_urbanictiy_7}
results_7 <- compute_urbanicity_batched(
  df          = reference_coordinate_set_7,
  ee_project  = ee_project_name,
  year_or_years = year_or_years,
  batch_size  = 5,
  save_prefix = paste0(output_fp, "subset7") 
)


```
