---
title: "Testing Different Composite Indices"
author: "Lev Kolinski"
date: "`r Sys.Date()`"
output: html_document
---

# Loading Packages
```{r setup}
library(tidyverse)
library(rnaturalearthdata)
library(rnaturalearth)
library(pecahnurbanicity)
```

# Loading In 1400 Urbanicity Reference Coordinates
```{r loading_reference}
ref_data <- readRDS("/Users/levkolinski/Desktop/pecahn-urbanicity/reference_data/urbanicity_reference_dataset.rds")

### getting lat/lon values
ref_data <- ref_data |>
  mutate(
    lat = as.numeric(gsub(",.*", "", community)),
    lon = as.numeric(gsub(".*, ", "", community))
  )
```

# Imputing weird nighttime light outlier
Subbing in the average nighttime light value for the other sites in that same smod class (22)

```{r impute_nightlight}
# get the community to replace
target_community <- ref_data |> 
  arrange(desc(nighttime_light_2025)) |> 
  slice(1) |> 
  pull(community)

# replace its nightlight values with smod_class 22 averages
years <- 2015:2025

avg_smod_11_nightlight <- sapply(years, function(yr) {
  col_name <- paste0("nighttime_light_", yr)
  ref_data |> 
    filter(smod_class == 22) |> 
    pull(all_of(col_name)) |> 
    mean(na.rm = TRUE)
})
names(avg_smod_11_nightlight) <- years

ref_data <- ref_data |> 
  mutate(across(
    all_of(paste0("nighttime_light_", years)),
    ~ if_else(community == target_community,
              avg_smod_11_nightlight[str_extract(cur_column(), "\\d{4}")],
              .x)
  ))

```

# Calculating Indices with Different Feature Sets

## All 6 Core Features
```{r all_features}
ref_data_comp_6 <- compute_urbanicity_composite_index(
  ref_data, method = "zscore",
  features = c(
    "travel_time_paved_road_min", "travel_time_hospital_min",
    "travel_time_school_min",     "travel_time_urban_center_min",
    "nighttime_light",            "pop_density"
  ),
  min_features = 6,
  suffix = "urban_index_6feat",
  keep_standardized = T
)
```

### Pearson Correlation Plot of 6 Core Features
```{r corr_plot}
z_features <- ref_data_comp_6 |> 
  select(travel_time_paved_road_min_z, travel_time_school_min_z, travel_time_urban_center_min_z, travel_time_hospital_min_z, nighttime_light_2025_z, pop_density_2025_z) |> 
  drop_na() |> 
  as.matrix()
  
M = cor(z_features,
                 method = "pearson")

corrplot::corrplot(M, method = "number",order = 'AOE')

```

## 4 Core Features: Nighttime Light, Population Density, Time to Urban Center, Time to Paved Road
```{r four_features}
ref_data_comp_4 <- compute_urbanicity_composite_index(
  ref_data, method = "zscore",
  features = c(
    "travel_time_urban_center_min", "nighttime_light",
    "pop_density",                  "travel_time_paved_road_min"
  ),
  min_features = 4,
  suffix = "urban_index_4feat",
  keep_standardized = T
)
```

## 3 Core Features: Nighttime Light, Population Density, Time to Urban Center
```{r three_features}
ref_data_comp_3 <- compute_urbanicity_composite_index(
  ref_data, method = "zscore",
  features = c(
    "travel_time_urban_center_min", "nighttime_light", "pop_density"
  ),
  min_features = 3,
  suffix = "urban_index_3feat",
  keep_standardized = T
)
```

# Comparing Composite Indices

```{r combine}
comparison <- data.frame(
  urban_index_6feat = ref_data_comp_6$urban_index_6feat,
  urban_index_4feat = ref_data_comp_4$urban_index_4feat,
  urban_index_3feat = ref_data_comp_3$urban_index_3feat
) |>
  mutate(
    rank_6 = rank(urban_index_6feat, na.last = "keep"),
    rank_4 = rank(urban_index_4feat, na.last = "keep"),
    rank_3 = rank(urban_index_3feat, na.last = "keep"),
    rank_shift_6v4 = (rank_6 - rank_4),
    rank_shift_6v3 = (rank_6 - rank_3),
    rank_shift_4v3 = (rank_4 - rank_3)
  )
```

# Feature Distributions

```{r feature_distributions, fig.width = 10, fig.height = 8}
feature_cols <- c(
  "travel_time_paved_road_min",
  "travel_time_hospital_min",
  "travel_time_school_min",
  "travel_time_urban_center_min",
  "nighttime_light_2025",
  "pop_density_2025", "pct_paved_roads"
)

feature_labels <- c(
  "travel_time_paved_road_min"  = "Travel time: paved road (min)",
  "travel_time_hospital_min"    = "Travel time: hospital (min)",
  "travel_time_school_min"      = "Travel time: school (min)",
  "travel_time_urban_center_min"= "Travel time: urban center (min)",
  "nighttime_light_2025"        = "Nighttime light (2025)",
  "pop_density_2025"            = "Population density (2025)",
   "pct_paved_roads"            = "% Paved Roads"

)

ref_data |>
  select(all_of(feature_cols)) |>
  pivot_longer(everything(), names_to = "feature", values_to = "value") |>
  mutate(feature = recode(feature, !!!feature_labels)) |>
  ggplot(aes(x = value)) +
  geom_histogram(bins = 60, fill = "#2166ac", alpha = 0.7) +
  facet_wrap(~ feature, scales = "free", ncol = 2) +
  labs(
    x     = NULL,
    y     = "Count",
    title = "Reference dataset: raw feature distributions (post-imputation)"
  ) +
  theme_minimal() +
  theme(strip.text = element_text(size = 9))
```

```{r feature_distributions_log, fig.width = 10, fig.height = 8}
ref_data |>
  select(all_of(feature_cols)) |>
  pivot_longer(everything(), names_to = "feature", values_to = "value") |>
  mutate(feature = recode(feature, !!!feature_labels)) |>
  ggplot(aes(x = value)) +
  geom_histogram(bins = 60, fill = "#d6604d", alpha = 0.7) +
  scale_x_continuous(trans = scales::log1p_trans())+
  facet_wrap(~ feature, scales = "free", ncol = 2) +
  labs(
    x     = "Value (log(1 + x) scale)",
    y     = "Count",
    title = "Reference dataset: raw feature distributions (log10 scale) (post-imputation)"
  ) +
  theme_minimal() +
  theme(strip.text = element_text(size = 9))
```

# Spearman Rank Correlations of Composite Indices
```{r correlations}
cor_matrix <- cor(
  comparison[, c("urban_index_6feat", "urban_index_4feat", "urban_index_3feat")],
  method = "spearman",
  use    = "pairwise.complete.obs"
)
round(cor_matrix, 3)
```

# Rank Shift Summaries
```{r rank_shifts}
cat("--- 6-feature vs. 4-feature ---\n")
print(summary(comparison$rank_shift_6v4))

cat("\n--- 6-feature vs. 3-feature ---\n")
print(summary(comparison$rank_shift_6v3))

cat("\n--- 4-feature vs. 3-feature ---\n")
print(summary(comparison$rank_shift_4v3))

cat("\nSites shifting > 100 ranks (6v4):", sum(comparison$rank_shift_6v4 > 100, na.rm = TRUE), "\n")
cat("Sites shifting > 100 ranks (6v3):", sum(comparison$rank_shift_6v3 > 100, na.rm = TRUE), "\n")
cat("Sites shifting > 100 ranks (4v3):", sum(comparison$rank_shift_4v3 > 100, na.rm = TRUE), "\n")
```

# Score Distributions
```{r distributions}
comparison |>
  select(urban_index_6feat, urban_index_4feat, urban_index_3feat) |>
  pivot_longer(everything(), names_to = "index", values_to = "score") |>
  mutate(index = recode(index,
    urban_index_6feat = "6 features",
    urban_index_4feat = "4 features",
    urban_index_3feat = "3 features"
  )) |>
  ggplot(aes(x = score, fill = index)) +
  geom_density(alpha = 0.4) +
  scale_fill_manual(values = c(
    "6 features" = "#2166ac",
    "4 features" = "#f4a582",
    "3 features" = "#d6604d"
  )) +
  labs(
    x = "Urbanicity index (average z-score)",
    y = "Density",
    fill = "Feature set",
    title = "Score distributions by feature set"
  ) +
  theme_minimal()
```

# Pairwise Scatter Plots
```{r scatterplots, fig.width = 9, fig.height = 3.5}
p1 <- ggplot(comparison, aes(x = urban_index_6feat, y = urban_index_4feat)) +
  geom_point(alpha = 0.3, size = 0.7) +
  geom_abline(slope = 1, intercept = 0, color = "red", linetype = "dashed") +
  labs(x = "6-feature index", y = "4-feature index",
       title = "6 vs. 4 features") +
  theme_minimal()

p2 <- ggplot(comparison, aes(x = urban_index_6feat, y = urban_index_3feat)) +
  geom_point(alpha = 0.3, size = 0.7) +
  geom_abline(slope = 1, intercept = 0, color = "red", linetype = "dashed") +
  labs(x = "6-feature index", y = "3-feature index",
       title = "6 vs. 3 features") +
  theme_minimal()

p3 <- ggplot(comparison, aes(x = urban_index_4feat, y = urban_index_3feat)) +
  geom_point(alpha = 0.3, size = 0.7) +
  geom_abline(slope = 1, intercept = 0, color = "red", linetype = "dashed") +
  labs(x = "4-feature index", y = "3-feature index",
       title = "4 vs. 3 features") +
  theme_minimal()

gridExtra::grid.arrange(p1, p2, p3, ncol = 3)
```

# Rank Shift Distributions
```{r rank_shift_distributions}
comparison |>
  select(rank_shift_6v4, rank_shift_6v3, rank_shift_4v3) |>
  pivot_longer(everything(), names_to = "comparison", values_to = "rank_shift") |>
  mutate(comparison = recode(comparison,
    rank_shift_6v4 = "6 vs. 4 features",
    rank_shift_6v3 = "6 vs. 3 features",
    rank_shift_4v3 = "4 vs. 3 features"
  )) |>
  ggplot(aes(x = rank_shift, fill = comparison)) +
  geom_histogram(binwidth = 25, alpha = 0.7, position = "identity") +
  scale_fill_manual(values = c(
    "6 vs. 4 features" = "#2166ac",
    "6 vs. 3 features" = "#d6604d",
    "4 vs. 3 features" = "#f4a582"
  )) +
  facet_wrap(~ comparison, ncol = 1) +
  labs(
    x    = "Absolute rank shift",
    y    = "Count",
    fill = "Comparison",
    title = "Rank shift distributions across feature sets"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```

# Largest Rank Movers (6-feature vs. 4-feature)
```{r big_movers_64}
# Sites with the largest rank shifts between full and second-smallest feature set
big_movers_64 <- comparison |>
  mutate(
    row_id = row_number(),
    lon    = ref_data$lon,
    lat    = ref_data$lat
  ) |>
  filter(abs(rank_shift_6v4) > 100) |>
  arrange(desc(abs(rank_shift_6v4))) |>
  select(row_id, lon, lat,
         urban_index_6feat, urban_index_4feat,
         rank_6, rank_4, rank_shift_6v4)

cat("Sites shifting > 100 ranks (6v4):", nrow(big_movers_64), "\n\n")
head(big_movers_64, 20)
```

```{r biggest_mover_map64}
world_map  <- ne_countries(scale = "medium", returnclass = "sf")

ggplot() +
  geom_sf(data = world_map, fill = "grey15", color = "grey35",
          linewidth = 0.15) +
  geom_point(
    data  = big_movers_64,
    aes(x = lon, y = lat, color =rank_shift_6v4),
    size = 1.3, alpha = 0.85, shape = 16
  ) +
  scale_color_viridis_c(
    option = "inferno",
    name   = "Rank\nshift"
  )+
  labs(
    title    = "Rank shift: 6 v 4 Feature",
    x = NULL, y = "% of global total"
  ) +
  coord_sf(expand = FALSE) +
  theme_void(base_size = 10) +
  theme(
    plot.background = element_rect(fill = "grey10", color = NA),
    plot.title      = element_text(color = "white", face = "bold", size = 11),
    plot.subtitle   = element_text(color = "grey70", size = 8),
    legend.position = "right",
    legend.title    = element_text(color = "white", size = 7.5),
    legend.text     = element_text(color = "grey85", size = 6.5),
    legend.key.size = unit(0.35, "cm"),
    plot.margin     = margin(5, 10, 8, 10)
  )
```


# Largest Rank Movers (6-feature vs. 3-feature)
```{r big_movers_63}
# Sites with the largest rank shifts between full and minimal feature set
big_movers <- comparison |>
  mutate(
    row_id = row_number(),
    lon    = ref_data$lon,
    lat    = ref_data$lat
  ) |>
  filter(abs(rank_shift_6v3) > 100) |>
  arrange(desc(abs(rank_shift_6v3))) |>
  select(row_id, lon, lat,
         urban_index_6feat, urban_index_3feat,
         rank_6, rank_3, rank_shift_6v3)

cat("Sites shifting > 100 ranks (6v3):", nrow(big_movers), "\n\n")
head(big_movers, 20)
```

```{r biggest_mover_map_63}
ggplot() +
  geom_sf(data = world_map, fill = "grey15", color = "grey35",
          linewidth = 0.15) +
  geom_point(
    data  = big_movers,
    aes(x = lon, y = lat, color =rank_shift_6v3),
    size = 1.3, alpha = 0.85, shape = 16
  ) +
  scale_color_viridis_c(
    option = "inferno",
    name   = "Rank\nshift"
  )+
  labs(
    title    = "Rank shift: 6 v 3 Feature",
    x = NULL, y = "% of global total"
  ) +
  coord_sf(expand = FALSE) +
  theme_void(base_size = 10) +
  theme(
    plot.background = element_rect(fill = "grey10", color = NA),
    plot.title      = element_text(color = "white", face = "bold", size = 11),
    plot.subtitle   = element_text(color = "grey70", size = 8),
    legend.position = "right",
    legend.title    = element_text(color = "white", size = 7.5),
    legend.text     = element_text(color = "grey85", size = 6.5),
    legend.key.size = unit(0.35, "cm"),
    plot.margin     = margin(5, 10, 8, 10)
  )
```

# Largest Rank Movers (4-feature vs. 3-feature)
```{r big_movers_43}
# Sites with the largest rank shifts between two smallest feature sets
big_movers_43 <- comparison |>
  mutate(
    row_id = row_number(),
    lon    = ref_data$lon,
    lat    = ref_data$lat
  ) |>
  filter(abs(rank_shift_4v3) > 100) |>
  arrange(desc(abs(rank_shift_4v3))) |>
  select(row_id, lon, lat,
         urban_index_4feat, urban_index_3feat,
         rank_4, rank_3, rank_shift_4v3)

cat("Sites shifting > 100 ranks (4v3):", nrow(big_movers_43), "\n\n")
head(big_movers_43, 20)
```

```{r biggest_mover_map_43}
ggplot() +
  geom_sf(data = world_map, fill = "grey15", color = "grey35",
          linewidth = 0.15) +
  geom_point(
    data  = big_movers_43,
    aes(x = lon, y = lat, color =rank_shift_4v3),
    size = 1.3, alpha = 0.85, shape = 16
  ) +
  scale_color_viridis_c(
    option = "inferno",
    name   = "Rank\nshift"
  )+
  labs(
    title    = "Rank shift: 4 v 3 Feature",
    x = NULL, y = "% of global total"
  ) +
  coord_sf(expand = FALSE) +
  theme_void(base_size = 10) +
  theme(
    plot.background = element_rect(fill = "grey10", color = NA),
    plot.title      = element_text(color = "white", face = "bold", size = 11),
    plot.subtitle   = element_text(color = "grey70", size = 8),
    legend.position = "right",
    legend.title    = element_text(color = "white", size = 7.5),
    legend.text     = element_text(color = "grey85", size = 6.5),
    legend.key.size = unit(0.35, "cm"),
    plot.margin     = margin(5, 10, 8, 10)
  )
```

# Rank Shifts by SMOD Class

```{r smod_rank_shifts}
smod_labels <- c(
  "11" = "Very Low Density Rural",
  "12" = "Low Density Rural",
  "13" = "Rural Cluster",
  "21" = "Suburban/Peri-Urban",
  "22" = "Semi-Dense Urban Cluster",
  "23" = "Dense Urban Cluster",
  "30" = "Urban Centre"
)

comparison_smod <- comparison |>
  mutate(
    row_id     = row_number(),
    smod_class = as.character(ref_data$smod_class),
    smod_label = recode(smod_class, !!!smod_labels),
    lon        = ref_data$lon,
    lat        = ref_data$lat
  )
```

## Mean Absolute Rank Shift by SMOD Class
```{r smod_mean_shift}
comparison_smod |>
  group_by(smod_class, smod_label) |>
  summarise(
    n               = n(),
    mean_shift_6v4  = mean(abs(rank_shift_6v4), na.rm = TRUE),
    mean_shift_6v3  = mean(abs(rank_shift_6v3), na.rm = TRUE),
    mean_shift_4v3  = mean(abs(rank_shift_4v3), na.rm = TRUE),
    .groups = "drop"
  ) |>
  arrange(as.numeric(smod_class))
```

## Rank Shift Distributions Faceted by SMOD Class
```{r smod_shift_boxplot, fig.width = 10, fig.height = 7}
comparison_smod |>
  select(smod_label, rank_shift_6v4, rank_shift_6v3, rank_shift_4v3) |>
  pivot_longer(
    cols      = starts_with("rank_shift"),
    names_to  = "comparison",
    values_to = "rank_shift"
  ) |>
  mutate(
    comparison = recode(comparison,
      rank_shift_6v4 = "6 vs. 4 features",
      rank_shift_6v3 = "6 vs. 3 features",
      rank_shift_4v3 = "4 vs. 3 features"
    ),
    smod_label = factor(smod_label, levels = smod_labels)
  ) |>
  ggplot(aes(x = smod_label, y = rank_shift, fill = comparison)) +
  geom_boxplot(alpha = 0.7, outlier.size = 0.5, outlier.alpha = 0.4) +
  scale_fill_manual(values = c(
    "6 vs. 4 features" = "#2166ac",
    "6 vs. 3 features" = "#d6604d",
    "4 vs. 3 features" = "#f4a582"
  )) +
  labs(
    x     = NULL,
    y     = "Absolute rank shift",
    fill  = "Comparison",
    title = "Rank shift by SMOD class and feature set"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 35, hjust = 1, size = 8),
    legend.position = "top"
  )
```

## Proportion of Sites with Large Rank Shifts (>100) by SMOD Class
```{r smod_large_shifts, fig.width = 10, fig.height = 5}
comparison_smod |>
  group_by(smod_class, smod_label) |>
  summarise(
    n                  = n(),
    pct_large_6v4      = mean(abs(rank_shift_6v4) > 100, na.rm = TRUE),
    pct_large_6v3      = mean(abs(rank_shift_6v3) > 100, na.rm = TRUE),
    pct_large_4v3      = mean(abs(rank_shift_4v3) > 100, na.rm = TRUE),
    .groups = "drop"
  ) |>
  pivot_longer(
    cols      = starts_with("pct_large"),
    names_to  = "comparison",
    values_to = "pct"
  ) |>
  mutate(
    comparison = recode(comparison,
      pct_large_6v4 = "6 vs. 4 features",
      pct_large_6v3 = "6 vs. 3 features",
      pct_large_4v3 = "4 vs. 3 features"
    ),
    smod_label = factor(smod_label, levels = smod_labels)
  ) |>
  ggplot(aes(x = smod_label, y = pct, fill = comparison)) +
  geom_col(position = "dodge", alpha = 0.85) +
  scale_y_continuous(labels = scales::percent) +
  scale_fill_manual(values = c(
    "6 vs. 4 features" = "#2166ac",
    "6 vs. 3 features" = "#d6604d",
    "4 vs. 3 features" = "#f4a582"
  )) +
  labs(
    x     = NULL,
    y     = "% of sites shifting > 100 ranks",
    fill  = "Comparison",
    title = "Proportion of sites with large rank shifts (> 100) by SMOD class"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 35, hjust = 1, size = 8),
    legend.position = "top"
  )
```
